Title

### 2015-07-22 main.tex
Figures included

% Package to generate and customize Algorithm as per ACM style
\usepackage[ruled]{algorithm2e}
\renewcommand{\algorithmcfname}{ALGORITHM}
\SetAlFnt{\small}
\SetAlCapFnt{\small}
\SetAlCapNameFnt{\small}
\SetAlCapHSkip{0pt}
\IncMargin{-\parindent}

@@

% Title portion
\title{Title}
\author{BONNIE BERGER
\affil{Massachusetts Institute of Technology}
\author{NOAH M. DANIELS
\affil{Massachusetts Institute of Technology}
\author{Y. WILLIAM YU
\affil{Massachusetts Institute of Technology}
% NOTE! Affiliations placed here should be for the institution where the
%       BULK of the research was done. If the author has gone to a new
%       institution, before publication, the (above) affiliation should NOT be changed.
%       The authors 'current' address may be given in the "Author's addresses:" block (below).
%       So for example, Mr. Abdelzaher, the bulk of the research was done at UIUC, and he is
%       currently affiliated with NASA.

@@

Certainly, computers are getting faster and more economical; the amount of 
processing available per dollar of compute hardware is more or less doubling 
every year or two; a similar claim can be made about storage capacity 
(Figure~\ref{moore}).
In 2002, when the first human genome was sequenced, the growth in computing 
power was still matching the growth rate of genomic data.
However, the sequencing technology used for the Human Genome Project--Sanger
sequencing--was supplanted around 2004, with the advent of what is now known as 
next-generation sequencing.
The material costs to sequence a genome have plummeted in the past
decade, to the point where a whole human genome can be sequenced for less than
one thousand U.S. Dollars.
As a result, by some estimates, the amount of genomic data available to 
researchers is increasing by a factor of 10 every year.

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have teamed up to bring the GATK (Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
ALSO MENTION AMAZON AWS HERE.
However, while cloud computing frees researchers from maintaining their own
data centers, and provides cost-saving benefits when computing resources are
not needed around the clock, it is no panacea.
First, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.
Moreover, cloud computing does not truly address the problem posed by
a corpus of data that is growing faster than Moore's law, because the computer
systems that make up those cloud datacenters are still themselves bound by
improvements in semiconductor technology; they still follow the trajectory
suggested by Moore's law.

cite nature article (Marks)
mention cloud computing, but it doesn't truly address the problem
algorithms for coming up with novel biological insights NOT focus of this review
storage requirements, network transmission, etc.

[image from: http://www.nature.com/nature/journal/v498/n7453/full/498255a.html; picture from EBI]

This Nature article pointed out that currently many omics applications require us to store, access and analyze large libraries of data.
However, the field is realizing that we don’t have the computational tools to tackle these tasks, as Arend Sidow noted.

It’s tempting to think that cloud computing is going to solve this problem.
But that’s not the case.
It might save a little money but doesn’t address the fundamental issue.
It doesn’t change the problem that the data is increasing exponentially faster than computing power per dollar.

Furthermore, as transmission bottlenecks grow as an issue, even getting data into the cloud presents a challenge.

The only solution is to discover fundamentally better algorithms for processing these databases.
Better algorithms can make an enormous difference.

In fact, we need [click] algorithms…
And that’s what we do.
We design algs that do these calculations a lot faster and space-efficiently, and thus their cost doesn’t explode as the size of the databases increase.
* In this talk, we will focus on sublinear-time algorithms, but we also get sublinear-space.

dim reduction - PCA, DCA
indexing - FM-index
kmer approaches (Kingsford)
compressed sensing (Linial)
low-dim polytopes (Alon)
tree decomposition
isorank

@@

Exploiting novel structure in biological data
- What data looks like
- actual fractal dimension & metric entropy of example data sets: table
 - metagenomics
 - protein seq
 - bacterial genomes
 - protein structure
 - chemogenomics
- Runtime
- Storage
- Search as exemplar.

@@

% Figure~\ref{fig:one} shows a typical microcontroller C program that
% controls an automotive power window lift. The program is one of the
% programs used in the case study described in Section~\ref{sec:sim}.
% At first sight, the programs looks like an ANSI~C program. It
% contains function calls, assignments, if clauses, and while loops.
% Figure
% \begin{figure}
% \centerline{\includegraphics{acmsmall-mouse}}
% \caption{Code before preprocessing.}
% \label{fig:one}
% \end{figure}

>>>>

\usepackage{caption}
\usepackage{subfig}

@@

% Title portion
\title{Title}
\author{BONNIE BERGER
\affil{Massachusetts Institute of Technology}
NOAH M. DANIELS
\affil{Massachusetts Institute of Technology}
Y. WILLIAM YU
\affil{Massachusetts Institute of Technology}}

@@

Certainly, computers are getting faster and more economical; the amount of 
processing available per dollar of compute hardware is more or less doubling 
every year or two; a similar claim can be made about storage capacity 
(Figure~\ref{fig:growth}).
In 2002, when the first human genome was sequenced, the growth in computing 
power was still matching the growth rate of genomic data.
However, the sequencing technology used for the Human Genome Project--Sanger
sequencing--was supplanted around 2004, with the advent of what is now known as 
next-generation sequencing.
The material costs to sequence a genome have plummeted in the past
decade, to the point where a whole human genome can be sequenced for less than
one thousand U.S. Dollars.
As a result, by some estimates, the amount of genomic data available to 
researchers is increasing by a factor of 10 every year.

\begin{figure}[htb!]
\centering
\subfloat[Growth of genomic sequence data as compared with the combined power of the top-500 supercomputer list; y-axis is logarithmic]{
\includegraphics[width=3.0in]{assets/moore.png}
\label{fig:moore}
}\\
\subfloat[Growth of genomic sequence data as compared with hard drive capacity; y-axis is logarithmic]{
\includegraphics[width=3.0in]{assets/kryder.png}
\label{fig:kryder}
}
\caption[Sequencing vs. Storage]{Moore's and Kryder's laws contrasted with genomic sequence data}
\label{fig:growth}
\end{figure}

This growth in data poses significant challenges for 
researchers~\cite{marx2013biology}.
Currently, many omics applications require us to store, access, and analyze
large libraries of data.
One approach to solving these challenges is to embrace cloud computing.
Google, Inc. and the Broad Institute have teamed up to bring the GATK (Genome 
Analysis Toolkit) to the Google cloud (https://cloud.google.com/genomics/gatk).
ALSO MENTION AMAZON AWS HERE.
However, while cloud computing frees researchers from maintaining their own
data centers, and provides cost-saving benefits when computing resources are
not needed around the clock, it is no panacea.
First, in the face of disease outbreaks such as the 2014 Ebola virus epidemic 
in West Africa, analysis resources are needed at often-remote field sites.
While it is now possible to bring sequencing equipment and limited computing 
resources to remote sites, internet connectivity is still highly constrained.
Thus, accessing cloud resources for analytics may not be possible.
Moreover, cloud computing does not truly address the problem posed by
a corpus of data that is growing faster than Moore's law, because the computer
systems that make up those cloud datacenters are still themselves bound by
improvements in semiconductor technology; they still follow the trajectory
suggested by Moore's law.

ALGORITHMS FOR COMING UP WITH NOVEL BIOLOGICAL INSIGHTS not FOCUS OF THIS REVIEW
STORAGE REQUIREMENTS, NETWORK TRANSMISSION, ETC.

Computer scientists routinely exploit peculiarities in the structure of data in
order to reduce time or space complexity.
In computational biology, this approach has served researchers well.
Now-classical approaches such as principal component analysis (PCA) reduce the 
dimensionality of data in order to simplify analysis and uncover salient 
features.

\section{Biological sequence data and next-generation sequencing}

Many, but not all, of the problems in computational biology deal with sequence
data, either nucleotide sequences (nominally a four-letter alphabet) or protein
sequences (nominally a twenty-letter alphabet).
Certainly, as sequencing is generating the greatest volume of biological data,
a large portion of bioinformatics algorithms deal primarily with string data.
For the problem of exact search, approaches for transforming and indexing 
string data provide benefits.
The Burrows-Wheeler transform (BWT)~\cite{burrows1994block} provides efficient string 
compression through a reversible transformation, while the FM-index data 
structure~\cite{ferragina2000opportunistic} is a compressed substring index, based
on the BWT, which provides efficient storage as well as fast search.
Modern next-generation sequencing (NGS) technologies read short fragments of
DNA, known as \emph{reads}~\ref{fig:shotgun}, which must be mapped onto a reference sequence~\ref{fig:ngs-pipeline}
BWA (Burrows-Wheeler Aligner)~\cite{li2009fast} uses the BWT, while the 
Bowtie2~\cite{langmead2012fast} aligner further relies on the FM-index for efficient 
mapping of NGS reads.

\begin{figure}[htb!]
\centering
\subfloat[``Shotgun'' sequencing breaks DNA molecules into many short fragments, or reads, and relies on high coverage to produce a statistically likely representation of a whole genome]{
\includegraphics[width=3.0in]{assets/shotgun.png}
\label{fig:shotgun}
}
\subfloat[Single-nucleotide polymorphisms, or SNPs, are the simplest type of genomic variant, and form the bulk of ``variant-calling'' analysis]{
\includegraphics[width=3.0in]{assets/snps.png}
\label{fig:snps}
}\\
\subfloat[The NGS downstream analysis pipeline. Shotgun reads are mapped to a reference genome with tools such as BWA or Bowtie. The resulting genomic sequence is analyzed for variants with tools such as GATK or Samtools. This allows relationships between genes and diseases to be uncovered.]{
\includegraphics[width=3.0in]{assets/ngs-pipeline.png}
\label{fig:ngs-pipeline}
}
\caption[The NGS Pipeline]{The next-generation sequencing (NGS) pipeline.}
\label{fig:ngs}
\end{figure}

As biological sequence data is string data, it is frequently useful to index or
count instances of $k$-mers of a fixed length, $k$.
Search tools such as BLAST~\cite{altschul1990basic} rely on $k$-mer \emph{seeds} in a 
seed-and-extend approach to sequence search and alignment.
For counting $k$-mer frequencies, Jellyfish~\cite{marccais2011fast} relies on an efficient
hash-table encoding as well as Intel's ``compare-and-swap'' CPU instructions for
improved run-time performance.

Seeking to take advantage of sequence redundancy, \cite{loh2012compressive}
introduced \emph{compressive genomics}, an approach that relies on compressing
data in such a way that the desired computation (such as BLAST search) can be
performed in the compressed representation.
This approach

(now, redundancy and compressive) loh2012compressive

@@

Quality score compression

lossy compression - quartz

Mince~\cite{patro2015data} groups similar reads (those that share a common 
substring) together into ``buckets'', allowing that common substring to now be 
removed and treated as the bucket label, so that each read in the compressed 
representation comprises only its unique differences from the bucket label.
This allows a general-purpose compressor to achieve tight compression.

\section{Non-sequence data}

Another important type of biological data is gene expression, or the quantity of
a particular gene product (usually a protein) present in the cell.
Gene expression data is useful for relating genotype to phenotype, and is 
important in studying diseases including cancer.
Usually obtained through microarrays or RNA-seq technologies, expression data is
quantitative, as each gene from a sample is associated with a numeric expression
level, and high-dimensional, as many thousands of genes may be analyzed at a
time.
While expression data lends itself to cluster analysis and probabilistic 
approaches, the high dimensionality presents challenges.
Principal Component Analysis has shown promise in reducing the dimensionality
of gene expression data~\cite{raychaudhuri2000principal, alter2000singular, yeung2001principal, misra2002interactive}.
[NEED CITATIONS FOR DIFFICULTIES -- HIGH DIM MICROARRAY DATA]
Some recent work has explored other ways to exploit the high-dimensional 
structure of the data.
SPARCLE (SPArse ReCovery of Linear combinations of Expression)~\cite{prat2011recovering}
brings ideas from compressed sensing~\cite{candes2005decoding, candes2006stable, donoho2006compressed} 
to gene expression analysis.
Specifically, SPARCLE relies on the idea that a randomly-chosen set of points 
are unlikely to be close in a lower-dimensional linear subspace.
Thus, SPARCLE approximates a gene's expression profile as a linear combination
of a few other genes' profiles, and then applies supervised learning to
predict relationships between genes.

Another recent and novel approach to exploiting the structure of gene expression
space is Parti (Pareto task inference)~\cite{hart2015inferring}, which describes a set of
data as a polytope, and infers the specific tasks represented by vertices of
that polytope from the features most highly enriched at those vertices.
In human breast tumors, and in mouse tissues, the expression data were
well-described by tetrahedra whose vertices were enriched for different tumor
types and biological functions, from which~\cite{hart2015inferring} infer four distinct
tasks.

Networks are frequently used to represent biological data, such as the genetic
and physical interactions among proteins, such as those in metabolic pathways.
Several approaches have taken advantage of the particular topologies of these
networks, such as diffusion-based approaches, which explore the topology of
networks through random walks.
IsoRank~\cite{singh2008global} and IsoRankN~\cite{liao2009isorankn} perform global multiple network 
alignment based on network topology and conserved biological function.
IsoRank's approach is similar to the PageRank algorithm~\cite{page1999pagerank}, while
IsoRankN produces a \emph{Personalized PageRank} alignment, similar to the
PageRank-Nibble algorithm~\cite{andersen2006local}.
In protein interaction networks, some nodes are high-degree \emph{hubs,} which 
interact with
many partners; these interactions do not imply shared function as strongly as
non-hub interactions.
Diffusion state distance (DSD)~\cite{cao2013going} is another diffusion-based approach
which defines a distance metric based on random walks.
DSD penalizes high-degree nodes, and outperforms other graph-theoretic distance
functions when applied to standard approaches such as neighbor voting for function annotation.

As new functional annotations and protein-protein interactions are discovered,
the number and size of these networks grows.
Diffusion component analysis (DCA)~\cite{cho} amounts to a network-based 
analogue of principle component analysis; DCA reduces the dimensionality of a
network in such a way that function annotation can be performed in the reduced 
space, significantly reducing running time while still providing competitive
prediction accuracy.

Finally, in the realm of protein structure prediction, it has been 
observed~\cite{dunbrack2002rotamer} that rather than exploring the continuous 
range of
possible dihedral angles, protein structures inhabit a smaller subspace, defined
by a narrower range of angles, referred to as 
rotamers~\cite{dunbrack1993backbone}.
The TreePack algorithm~\cite{xu2006fast} takes advantage of this observation 
for the
purpose of sidechain packing (finding the lowest-energy conformation for the
sidechain of each amino acid), using tree width to reduce the search space and
simplify dependencies.

Exploiting novel structure in biological data
- What data looks like
- actual fractal dimension and metric entropy of example data sets: table
 - metagenomics
 - protein seq
 - bacterial genomes
 - protein structure
 - chemogenomics
- Runtime
- Storage
- Search as exemplar.

@@
